---
title: "Homework 2"
format: html
editor: visual
author: Shey Dorji, Alecia Dezzani
theme: lux
---

## Getting Started: Data & Packages

```{r}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(jtools)
library(knitr)
library(kableExtra)
library(janitor)
library(geomtextpath)
```

```{r}
#install.packages("dataRetrieval")
#library(dataRetrieval)

#commenting out loading the data from NWIS 
#siteNumbers <- c("14211720", "14211010", "14166000")
#parameterCd <- "00010"  # water temperature
#startDate <- "2010-01-01"
#endDate <- "2023-12-31"

#temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

```

#### Generating Data
```{r}
# rename columns for clarity
#colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

# extract some date info for use later
#temp_data$month = month(temp_data$Date, label=TRUE)
#temp_data$year = year(temp_data$Date)

# check by plotting
#ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  #geom_line() +
  #labs(title="Stream Temperature at Willamette River Sites",
      # x="Date", y="Temperature (°C)", color="Site Number") +
  #theme_minimal()
```

#### Save & Load Data Frame
```{r}
#save the dataframe
#write_csv(temp_data, here("temp_data.csv"))

#local data load
temp_data_final <- read.csv(here("temp_data_final.csv"))
```

## Plot All Water Temperature Data
```{r data plot}

ggplot(temp_data_final, aes(x = month, y = year, fill = temp_C)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_distiller(palette = "Blues", name = "Temp (C)") +
  labs(
    title = "Monthly Temperature Heatmap",
    x = "Month",
    y = "Year",
    fill = "Temperature (C)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## Part I: Analysis of **August** Water Temperatures Over Time

### a) Is there a trend in temperature in the month of August?
```{r}
# get data for August
aug_data <- temp_data_final |> 
  clean_names() |>
  filter(month == "Aug")

# is there a trend between year and temperature? Is the temperature changing over time?
aug_lm <- lm(temp_c ~ year, data = aug_data)
summ(aug_lm, digits = 4)
```

Yes, there is a significant relationship between the year and the water temperature in August.
The p-value of the coefficient `year` is 0.0000 which is less than our alpha threshold of 0.05.

### b) Is the trend positive (increasing) or negative (decreasing)?
```{r}
aug_plot <- ggplot(aug_data, aes(x = year, y = temp_c, color = as.factor(year))) +
  theme_bw() +
  geom_jitter() +
  geom_smooth(aes(group = 1),
              method = "lm",
              se = FALSE,
              color = "black") +
  labs(title = "August Recorded Water Temperature for all sites",
       x = "Year",
       y = "Temp (deg C)",
       color = "Year")
  
aug_plot
```

Based on the plot above and the regression coefficient = 0.1512, the linear regression estimates that the August water temperature is increasing by 0.1512 degrees Celsius every year. However, this linear model has an R^2^ = 0.0804; therefore, only 8% of the variation in temperature can be attributed to a difference in year. Therefore, the year may not be the best tool to predict change in water temperature. 

### c) Are there different trends in each site?
```{r}
# OLS regression for Site 14211720
aug_1720 <- aug_data |>
  filter(site_no == 14211720)

aug_1720_lm <- lm(temp_c ~ year, 
                  data = aug_1720)
summ(aug_1720_lm, digits = 4)
```
```{r}
# OLS regression for Site 14211010
aug_1010 <- aug_data |>
  filter(site_no == 14211010)

aug_1010_lm <- lm(temp_c ~ year, 
                  data = aug_1010)
summ(aug_1010_lm, digits = 4)
```
```{r}
# OLS regression for Site 14166000
aug_6000 <- aug_data |>
  filter(site_no == 14166000)

aug_6000_lm <- lm(temp_c ~ year, 
                  data = aug_6000)
summ(aug_6000_lm, digits = 4)
```

```{r}
aug_site_plot <- ggplot(aug_data, aes(x = year, 
                                      y = temp_c, 
                                      color = as.factor(site_no))) +
  theme_bw() +
  geom_point(size = 2, alpha = 0.5) + 
  geom_labelsmooth(aes(label = site_no), 
                   fill = "white",
                method = "lm", 
                formula = y ~ x,
                size = 2.75, 
                linewidth = 1, 
                boxlinewidth = 0.4) +
  labs(title = "August Water Temperature Trends Per Site",
       x = "Year",
       y = "Temp (degrees C)",
       color = "Site") +
  guides(color = 'none')

aug_site_plot
```
**Yes**, there are differences in temperature trends between sites. At site 14211720, the linear model estimates that the water increases by 0.1459 degrees C per year. At site 14211010, the linear model estimates that the water increases by 0.1760 degrees C per year. At site 14166000, the linear model estimates that the water increases by 0.1297 degrees C per year. The figure above (*"August Water Temperature Trends Per Site"*) shows that each site has its own temperature trends. These results make sense, as each site likely has its own unique combination of upstream water flow, water depth, sun exposure, flow speed, et cetera. 
