---
title: "hw-2"
format: html
editor: visual
---

## Running Code

```{r}
library(tidyverse)
library(lubridate)
library(here)
```

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
#install.packages("dataRetrieval")
library(dataRetrieval)

#commenting out loading the data from NWIS and replacing with local csv load (Shey)

#siteNumbers <- c("14211720", "14211010", "14166000")
#parameterCd <- "00010"  # water temperature
#startDate <- "2010-01-01"
#endDate <- "2023-12-31"

#temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

temp_data <- read.csv(here("temp_data.csv"))

```

```{r}
# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)


# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

```{r}
write_csv(temp_data, here("temp_data.csv"))
```


1in your quarto file, add the option to load the data instead of accessing from NWIS server (completed -Shey)

2add a code chunk that plots the data (e.g. using ggplot) 

3add a code chunk that calculates some summary statistics (e.g. mean, max, min temperature for each site)
