---
title: "Homework 2"
format: html
editor: visual
author: Shey Dorji, Alecia Dezzani
theme: lux
---

## Getting Started: Access Data & Packages

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(jtools)
library(knitr)
library(kableExtra)
library(janitor)
library(geomtextpath)
```

```{r}
#install.packages("dataRetrieval")
#library(dataRetrieval)

#commenting out loading the data from NWIS 
#siteNumbers <- c("14211720", "14211010", "14166000")
#parameterCd <- "00010"  # water temperature
#startDate <- "2010-01-01"
#endDate <- "2023-12-31"

#temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

```

#### Generating Data
```{r}
# rename columns for clarity
#colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

# extract some date info for use later
#temp_data$month = month(temp_data$Date, label=TRUE)
#temp_data$year = year(temp_data$Date)

# check by plotting
#ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  #geom_line() +
  #labs(title="Stream Temperature at Willamette River Sites",
      # x="Date", y="Temperature (°C)", color="Site Number") +
  #theme_minimal()
```

#### Save & Load Data Frame
```{r}
#save the dataframe
#write_csv(temp_data, here("temp_data.csv"))

#local data load
temp_data_final <- read.csv(here("temp_data_final.csv"))
```

## Plot Water Temperature Data
```{r data plot}

ggplot(temp_data_final, aes(x = month, y = year, fill = temp_C)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_distiller(palette = "Blues", name = "Temp (C)") +
  labs(
    title = "Monthly Temperature Heatmap",
    x = "Month",
    y = "Year",
    fill = "Temperature (C)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## Part I: Analysis of August Water Temperatures

#### a) Is there a trend in temperature in the month of August?
```{r}
# get data for August
aug_data <- temp_data_final |> 
  clean_names() |>
  filter(month == "Aug")

# is there a trend between year and temperature? Is the temperature changing over time?
aug_lm <- lm(temp_c ~ year, data = aug_data)
summ(aug_lm, digits = 4)
```

Yes, there is a significant relationship between the year and the water temperature in August.
The p-value of the coefficient `year` is 0.0000 which is less than our alpha threshold of 0.05.

#### b) Is the trend positive (increasing) or negative (decreasing)?
```{r}
aug_plot <- ggplot(aug_data, aes(x = year, y = temp_c, color = as.factor(year))) +
  theme_bw() +
  geom_jitter() +
  geom_smooth(aes(group = 1),
              method = "lm",
              se = FALSE,
              color = "black") +
  labs(title = "August Recorded Water Temperature for all sites",
       x = "Year",
       y = "Temp (deg C)",
       color = "Year")
  
aug_plot
```

Based on the plot above and the regression coefficient = 0.1512, the linear regression estimates that the August water temperature is increasing by 0.1512 degrees Celsius every year. However, this linear model has an R^2^ = 0.0804 and the figure shows clearly that the data is not obviously linear. Therefore, only 8% of the variation in temperature can be attributed to a difference in year, and year is not be the best variable to estimate or predict change in water temperature. 

#### c) Are there different trends in each site?
```{r}
# OLS regression for Site 14211720
aug_1720 <- aug_data |>
  filter(site_no == 14211720)

aug_1720_lm <- lm(temp_c ~ year, 
                  data = aug_1720)
summ(aug_1720_lm, digits = 4, model.fit = FALSE, model.info = FALSE)
```
```{r}
# OLS regression for Site 14211010
aug_1010 <- aug_data |>
  filter(site_no == 14211010)

aug_1010_lm <- lm(temp_c ~ year, 
                  data = aug_1010)
summ(aug_1010_lm, digits = 4, model.fit = FALSE, model.info = FALSE)
```
```{r}
# OLS regression for Site 14166000
aug_6000 <- aug_data |>
  filter(site_no == 14166000)

aug_6000_lm <- lm(temp_c ~ year, 
                  data = aug_6000)
summ(aug_6000_lm, digits = 4, model.fit = FALSE, model.info = FALSE)
```

```{r}
aug_site_plot <- ggplot(aug_data, aes(x = year, 
                                      y = temp_c, 
                                      color = as.factor(site_no))) +
  theme_bw() +
  geom_point(size = 2, alpha = 0.5) + 
  geom_labelsmooth(aes(label = site_no), 
                   fill = "white",
                method = "lm", 
                formula = y ~ x,
                size = 2.75, 
                linewidth = 1, 
                boxlinewidth = 0.4) +
  labs(title = "August Water Temperature Trends Per Site",
       x = "Year",
       y = "Temp (degrees C)",
       color = "Site") +
  guides(color = 'none')

aug_site_plot
```
**Yes**, there are differences in temperature trends between sites. At site 14211720, the linear model estimates that the water increases by 0.1459 degrees C per year. At site 14211010, the linear model estimates that the water increases by 0.1760 degrees C per year. At site 14166000, the linear model estimates that the water increases by 0.1297 degrees C per year. The figure above (*"August Water Temperature Trends Per Site"*) shows that each site has its own temperature trends. These results make sense, as each site likely has its own unique combination of upstream water flow, water depth, sun exposure, flow speed, et cetera. 

## Management Recommendations: Seasonal Water Temperature Changes

How large are the seasonal temperature differences, and do they change over time?
```{r}
#|message: false

#get data for August and December
summary_temp_data <- temp_data_final |>
  clean_names() |>
  filter(month %in% c("Aug", "Dec")) |>
  select(site_no, date, temp_c, month, year) 

#calculate mean water temp per month per year 
mean_temps <- summary_temp_data |>
  group_by(year, month) |>
  summarise(
    mean_temp = mean(temp_c),
  ) 
mean_temps

diff_temps <- mean_temps|>
  pivot_wider(names_from = month, values_from = mean_temp) |>
  clean_names() |>
  mutate(temp_diff = aug - dec,
         temp_diff = round(temp_diff, 2))
diff_temps

min(diff_temps$temp_diff) # equals 12.47
mean(diff_temps$temp_diff) # equals 13.90
max(diff_temps$temp_diff) # equals 15.42
```

```{r}
# Plot seasonal temperature means

mean_temps_plot <- ggplot(mean_temps, aes(x = year,
                                          y = mean_temp,
                                          color = as.factor(year))) + 
  stat_summary(fun.min = min, fun.max = max, geom = "linerange", color = "grey") + 
  geom_point(aes(shape = as.factor(month)), size = 2.5) +
  labs(x = "Year",
       y = "Temperature (C)",
       title = "(A) Mean Monthly Water Temperature (August and December)",
       color = "Year",
       shape = "Month") +
  theme_minimal()
mean_temps_plot
```

```{r}
# plot annual difference in seasonal temperatures
diff_plot <- ggplot(diff_temps, aes(x = year, y = temp_diff)) +
  geom_bar(stat = "identity", width = 0.7, aes(fill = as.factor(year))) +
  geom_text(aes(label = round(temp_diff, 2), vjust = -0.5), size = 2.8)+
  labs(x = "Year",
       y = "Difference in Aug-Dec Temperature (C)",
       title = "(B) Seasonal Difference in Water Temperature") +
  guides(fill = "none") +
  theme_minimal()
diff_plot
```
```{r}
library(patchwork)
mean_temps_plot + diff_plot
```
Figure A above shows the difference in mean water temperature between August and December annually from 2010 through 2023. The seasonal temperature difference between August and December ranges from 12.47 deg C to 15.42 deg C, with a mean of 13.90 deg C difference. For conservation planning purposes, it is reasonable to estimate that the water temperature will decrease approximately 13-14 degrees between the summer and the winter this season. In addition, Figure B plots the water temperature difference over time. There are no clear outliers, and the distribution is relatively uniform. Therefore, it is expected that near-term seasonal water temperature decreases will follow a similar pattern. Even though the analysis of August and December data shows that the temperatures are increasing each year, they appear to be increasing at a consistent rate so that the seasonal change stays relatively constant. 


