---
title: "Homework 2"
format: html
editor: visual
author: Shey Dorji, Alecia Dezzani
theme: lux
---

## Getting Started: Data & Packages

```{r}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(jtools)
library(knitr)
library(kableExtra)
library(janitor)
library(geomtextpath)
```



```{r}
#install.packages("dataRetrieval")
library(dataRetrieval)

#commenting out loading the data from NWIS 
#siteNumbers <- c("14211720", "14211010", "14166000")
#parameterCd <- "00010"  # water temperature
#startDate <- "2010-01-01"
#endDate <- "2023-12-31"

#temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

```

```{r}
# rename columns for clarity
#colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

# extract some date info for use later
#temp_data$month = month(temp_data$Date, label=TRUE)
#temp_data$year = year(temp_data$Date)

# check by plotting
#ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  #geom_line() +
  #labs(title="Stream Temperature at Willamette River Sites",
      # x="Date", y="Temperature (°C)", color="Site Number") +
  #theme_minimal()
```

```{r}
#save the dataframe
#write_csv(temp_data, here("temp_data.csv"))

#local data load
temp_data <- read.csv(here("temp_data.csv"))
```

```{r data plot}

#2. Plot the data

ggplot(temp_data, aes(x = month, y = year, fill = temp_C)summ) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_distiller(palette = "Blues", name = "Temp (C)") +
  labs(
    title = "Monthly Temperature Heatmap",
    x = "Month",
    y = "Year",
    fill = "Temperature (C)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r}
#3. Summary statistics (e.g. mean, max, min temperature for each site)

temp_summary_stats<- temp_data |> group_by(site_no) |> summarize(mean_temp = mean(temp_C), max_temp = max(temp_C), min_temp = min(temp_C)) |> ungroup()


kable(temp_summary_stats)

```

## August Analysis

### a) Is there a trend in temperature in the month of August?
```{r}
# get data for August
aug_data <- temp_data |> 
  clean_names() |>
  filter(month == "Aug")

# is there a trend between year and temperature? Is the temperature changing over time?
aug_lm <- lm(temp_c ~ year, data = aug_data)
summ(aug_lm, digits = 4)

```
Yes, there is a significant relationship between the year and the water temperature in August.
The p-value of the coefficient `year` is 0.0000 which is less than our alpha threshold of 0.05.

### b) Is the trend positive (increasing) or negative (decreasing)?
```{r}
aug_plot <- ggplot(aug_data, aes(x = year, y = temp_c, color = as.factor(year))) +
  theme_bw() +
  geom_jitter() +
  geom_smooth(aes(group = 1),
              method = "lm",
              se = FALSE,
              color = "black") +
  labs(title = "August Recorded Water Temperature for all sites",
       x = "Year",
       y = "Temp (deg C)",
       color = "Year")
  
aug_plot
```
Based on the plot above and the regression coefficient = 0.1512, the linear regression estimates that the August water temperature is increasing by 0.1512 degrees Celcius every year. However, this linear model has an R^2^ = 0.0804; therefore, only 8% of the variation in temperature can be attributed to a difference in year. Therefore, the year may not be the best tool to predict change in water temperature. 

### c) Are there different trends in each site?
```{r}
aug_1720 <- aug_data |>
  filter(site_no == 14211720)

aug_1720_lm <- lm(temp_c ~ year, 
                  data = aug_1720)
summ(aug_1720_lm, digits = 4)
```


```{r}
aug_site_plot <- ggplot(aug_data, aes(x = year, y = temp_c, color = as.factor(site_no))) +
  theme_bw() +
  geom_point() + 
  geom_labelsmooth(aes(label = site_no), 
                   fill = "white",
                method = "lm", 
                formula = y ~ x,
                size = 2, 
                linewidth = 1, 
                boxlinewidth = 0.4) +
  labs(title = "August Recorded Water Temperature (deg C)",
       x = "Year",
       y = "Temp (deg C)",
       color = "Site") +
  guides(color = 'none')

aug_site_plot
```

